---
title: "1. R language basics, installing packages and RMarkdown usage"
author: "Ana J. Alegre (jalegre@centrogeo.edu.mx), Cristian Silva (csilva@centrogeo.edu.mx)"
output: html_notebook
date: "11/3/2021"
---

## Introduction

R is a programming language mainly aimed to statistical computing. It is a free software widely used for developing data analysis and more recently, for modern applications such as data science, machine learning or artificial intelligence.

In the field of geocomputation, R can be used for analyze, process, model and visualize geographic data by using the add-on packages available which extends the capabilities of the base language, enabling the user to use it as an effective Geographic Information System (GIS).

## Objective

The objective of this section is review the basic functions of R language in order to get familiar with its environment and syntax, install add-on packages to be used during the workshop and provide examples by doing through interactive RMarkdown notebooks.

## Base R review

### Getting help

To get help of a particular function, use `help`:
```{r Help of a function}
help("for") # Also ?for can be used on R console or highlight the command and pressing F1 on RStudio
```

Search a term on the help files using `help.search`:
```{r Search in help}
help.search("control flow")
```

### Basic operators

Example of addition:
```{r Basic operators}
34 + 56 # Addition
45 - 13 # Substract
34 * 0.14 # Multiply
89.4 / 5 # Divide
```


### Variable assignation

Assign a value to a variable with `<-` or `=`:
```{r Numeric variable assign}
my_number = 45
my_number # Visualize variable value
```

```{r String variable assign}
my_string <- "Hello world"
my_string # Visualize variable value
```

### Data types

Similar to other programming languages, R handles different basic data types for storing data. It is important to understand these data types in order to use R functions correctly.

#### Numeric

Numbers that have a decimal value or are a fraction:
```{r Numeric data type example}
a <- 1.5

# Previsualize a value:
a
```

To check the data type, use the `class` function:
```{r Check data type}
class(a)
```

#### Integer

Numbers that do not contain decimal values have a data type as an integer. By default if you write a number without decimal value R interprets it as numeric, so it is necessary to explicitly declare it as integer using the `as.integer` function:
```{r Integer data type}
a <- as.integer(1.5) # When casting the number to integer, it will truncate decimal value
a
```

Now, check the new data type of `a` variable:
```{r Check integer data type}
class(a)
```

#### Character

It can be a letter or a combination of letters enclosed by quotes is considered as a character data type by R. It can be alphabets or numbers.
```{r Character data type}
a <- "I like this workshop! (we hope so)" # When casting the number to integer, it will truncate decimal value
a
```

Now, check the new data type of `a` variable:
```{r Check character data type}
class(a)
```

The character data types are also known as *strings*.

#### Logical

A variable that can only have a value of `TRUE` and `FALSE` like a boolean is called a *logical variable*:
```{r Logical data type}
a <- TRUE
a
```

Now, check the new data type of `a` variable:
```{r Check logical data type}
class(a)
```

#### Vectors

Vectors are basic data structures used in R to compute multiple values at once and stores elements with the same data type . Create a vector using `c()`:
```{r Create vector}
my_vector <- c(5, 2, 4, 9, 8, 9, 1, 2, 2)
my_vector
```

Now, check the new data type of `my_vector` variable:
```{r Check vector elements data type}
class(my_vector) # Data type of the elements of the vector
```

#### Factor

They are used to refer to a qualitative relationship like colors, good & bad, course or movie ratings, etc. They are useful in statistical modeling. To define a factor, use a vector with the elements to be used and the function `factor`:

```{r Create a factor}
a <- factor(c("low", "low", "high", "medium", "high", "medium", "low"))
a
```

To check the levels of a factor, use the `levels` function:
```{r Check levels of a factor}
levels(a)
```

Now, check the new data type of `a` variable:
```{r Check factor data type}
class(a)
```

### Operations with vectors

Create a simple integer sequence:
```{r Create sequence}
2:10
```

Create a complex sequence using `seq`:
```{r Complex sequence vector}
seq(from = 0, to = 10, by = 2.5)
```

Repeat a vector using `rep`:
```{r Repeat vector}
rep(1:3, times = 4) # Repeat vector n times
rep(1:2, each = 3) # Repeat n times each value of the vector
```

Sort a vector values using `sort`:
```{r Sort vector}
sort(my_vector)
sort(my_vector, decreasing = TRUE) # Decreasing order
```

Return reversed vector using `rev`:
```{r Reverse vector}
rev(my_vector)
```

Return count of values using `table`:
```{r Count of vector values}
table(my_vector)
```

Return unique values of the vector using `unique`:
```{r Unique values in vector}
unique(my_vector)
```

Select a particular element in a vector (indexing starts from 1):
```{r Select by index}
my_vector[4] # Select the 4th element in my_vector
```

Select all elements but a particular one:
```{r Select but index}
my_vector[-4] # Select all but the 4th element in my_vector
```

Select a range of elements of a vector:
```{r Select by range}
my_vector[2:5] # Select the 2nd to 5th element in my_vector
```

Select all elements but a range:
```{r Select but range}
my_vector[-(2:5)] # Select all elements but the 2nd to 5th in my_vector
```

Select particular elements:
```{r Select by particular indexes}
my_vector[c(1,5)] # Select only 1st and 5th elements in my_vector
```

Select elements by value:
```{r Select by value}
my_vector[my_vector == 2] # Select elements equal to 2 in my_vector
```

Select elements by condition:
```{r Select by condition}
my_vector[my_vector < 3] # Select elements less than 3 in my_vector
```

Select elements by particular values:
```{r Select particular indexes}
my_vector[my_vector %in% c(1, 2, 5)]
```

### Conditions

Assign two variables:
```{r Variables to compare}
a <- 5
6 -> b # Reversed assign
```

Compare `a` and `b`:
```{r Conditions with variables}
a == b # Equal
a != b # Not equal
a > b # a greater than b
a >= b # a greater than or equal to b
a < b # a less than b
a <= b # a less than or equal than b
is.na(a) # Is missing (NA)
is.null(b) # Is null (NULL)
```

Comparing vectors:
```{r Conditions with vectors}
my_vector > a # Compare each element in my_vector to a
b %in% my_vector # Is b in any element of my_vector?
```

### Data frames

In R, a dataframe is a tightly coupled collection of variables built as a table where every column has a particular data type. Data frames are the fundamental data structure used by most of R modeling software. Create a data frame using the `data.frame` function:
```{r Create a data frame}
my_df <- data.frame(x = 1:5, y = c("A", "B", "C", "D", "E"))
my_df 
```

It is possible to browse the full contents of the data frame using the `View` function (in RStudio, a new window will open):
```{r View data frame}
View(my_df)
```

If just a preview of the data frame is needed, use the `head` function to visualize the *n* first rows:
```{r Head data frame}
head(my_df, n = 3) # If n parameter is ommited, head will show the first 5 rows by default
```

You can also browse the dataframes using the "Environment" panel in RStudio.

To access a particular row in a data frame, use the index between brackets ([Row, Column]):
```{r Access row in data frame}
my_df[1, ] # Omit column index to access entire row
```

To access a particular column in a data frame, use this syntax:
```{r Access column in data frame by index}
my_df[, 1] # Omit row index to access entire column values
```

The most common way to access values in a variable (column) is using the `$` separator:
```{r Access column by variable name}
my_df$y # Using variable name
```

Subset a particular element in a data frame using indexes:
```{r Subset element in data frame}
my_df[2, 2] # Access row 2, column 2
```

Understand the data frame dimensions with these functions:
```{r Data set dimensions}
dim(my_df) # Dimensions of data frame (row and column numbers)
nrow(my_df) # Number of rows in data frame
ncol(my_df) # Number of columns in data frame
```

Add a column using `cbind` function:
```{r Add column to data frame}
new_column <- data.frame(z = c(6:10)) # Must have the same rows than data frame
my_df <- cbind(my_df, new_column)
my_df
```

Add a row using `rbind` function:
```{r Add row to data frame}
new_row <- my_df[2, ] # Add a copy of row 2, must have same variable names else will add new columns with remaining names
my_df <- rbind(my_df, new_row)
my_df
```

### Programming

Repeat instructions in a loop using `for`:
```{r For loop}
for(i in 1:10) {
  j <- i * 10
  print(j)
}
```

Repeat instructions in a condition-controlled loop using `while` (if condition is not well defined, it may loop infinitely):
```{r While loop}
i <- 1
while(i < 10) {
  print(i)
  i <- i + 1
}
```

Run a code block when a condition is true, or optionally another code block when not with `ìf`:
```{r If sentence}
k <- 14
if(k > 10) {
  print("Variable k is greater than 10")
} else {
  print("Variable k is less or equal than 10")
}
```

Create a function to reuse a code block multiple times using `function`:
```{r Create function}
calculate_percent <- function(part, total) {
  percent <- part / total * 100
  return(percent)
}

# Call the function:
calculate_percent(part = 15, total = 234)
calculate_percent(15, 234) # Bypassing parameter names
```

### Working directory

As a best practice, it is recommended to set a working directory to store scripts, notebooks and data for better organization and easier coding. Get the current working directory using `getwd` function:
```{r Get working directory}
getwd()
```

Set the working directory using the `setwd` function, write between quotes and autocomplete pressing Tab key to set the path to your downloaded workshop folder:
```{r Set working directory}
# setwd("C:\iGISc") # Set path example on Windows
# setwd("/Users/username/Documents/iGISc") # Set path example on macOS or Linux
```

Also, you can use projects in RStudio to set the working directory to the folder you are working in.

### Read and write data

R uses a special format with extension `.RData` to store data used during a session, save this session variables in a file:
```{r Save variables as RData}
save(my_vector, my_string, my_df, file = "Data/my_data.RData")
```

Load variables stored as `.RData` using function `load`:
```{r Read variables from RData}
load("Data/my_data.RData")
```

R also can save data in plain text formats such as delimited or standard CSV (comma separated values), using these functions:
```{r Write data as plain text}
write.table(my_df, file = "Data/my_df.txt", sep = "|") # Save data frame as delimited text using "|" as separator
write.csv(my_df, file = "Data/my_df.csv") # Save data frame as CSV using "," as separator
write.csv2(my_df, file = "Data/my_df2.csv") # Save data frame as CSV using ";" as separator (for countries where decimal point is ",")
```

Read the data stored as plain text and store it a new variable using these functions:
```{r Read data from plain text}
my_df_from_delimited <- read.csv("Data/my_df.txt", sep = "|")
my_df_from_csv <- read.csv("Data/my_df.csv")
my_df_from_csv2 <- read.csv2("Data/my_df2.csv")

# Visualize loaded datasets:
my_df_from_delimited
my_df_from_csv
my_df_from_csv2
```

### Environment

List all variables in the environment using `ls`:
```{r List all variables}
ls()
```

Remove a variable in the environment using `rm`:
```{r Remove a variable}
rm("my_df") # Remove a variable
rm(list = ls()) # Remove all variables
```

You can also use the "Environment" panel in RStudio to browse and remove variables in your environment.

## Installing packages

R language extends its capabilities by using third-party packages. These packages are properly stored and documented on the [Comprehensive R Archive Network](https://cran.r-project.org) (aka. CRAN), allowing the users to add new functions to the ones includes on base R.

In this section you will be introduced to the functionality of several packages to be used during the workshop, commonly used to read, write, clean, manipulate, process and visualize data.

To install a particular package from CRAN, use `install.packages`:
```{r Install a package}
install.packages("carData") # carData is a package containing sample data
```

Another option to load, install and update packages is using the "Packages" panel on RStudio. 

Load a package for use in a session using `library` function:
```{r Load a package}
library(carData)
```

Use a particular function from a package without loading it writing the package name, double colon and then the desired function (the package must be already installed):
```{r Call a particular function from package}
utils::sessionInfo() # Use the sessionInfo function from utils package, browse available functions writing utils:: and pressing Tab key
```

Some packages includes sample data which can be loaded using the `data` function:
```{r Load sample data from package}
#carData:: # Press Tab key to browse the sample datasets included in the package
data(Salaries, package = "carData") # Load the Salaries dataset from package carData and store it as a variable
Salaries # Visualize dataset
```

## Commonly used packages

### Tidyverse

The `tidyverse` package is a collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures. Tidyverse includes, but is not limited to these packages:

* `tibble`: enables using an enhanced data structure format.
* `ggplot2`: for creating charts and advanced visualizations.
* `dplyr`: for data wrangling.
* `tidyr`: for data reshaping.
* `forcats`: enhances factors handling.
* `stringr`: better string manipulation.
* `readr`: for reading several data file formats.
* `purrr`: for better functional programming.

There are more packages sharing the `tidyverse` paradigm, its usage is often considered for state-of-the-art and modern R programming. 

To install `tidyverse` package:
```{r Install tidyverse}
install.packages("tidyverse")
```

For further reference, visit the [tidyverse website](https://www.tidyverse.org) and the [official package documentation](https://cran.r-project.org/web/packages/tidyverse/).

**Note for Linux users**: When installing `tidyverse` for the first time on Linux, it may be needed to install previously some libraries using the Terminal. Run the next command if using Debian-based Linux and then try again installing `tidyverse`:
```{bash Install Linux libraries for tidyverse}
sudo apt install libcurl4-openssl-dev libssl-dev libxml2-dev
```

### Lubridate

The `lubridate` package is used to make easier working with date-time formats and enable some new time-related functions in R. To install `lubridate` package:
```{r Install lubridate package}
install.packages("lubridate")
```

For further reference, visit the [lubridate website](https://lubridate.tidyverse.org) and the [official package documentation](https://cran.r-project.org/web/packages/lubridate/index.html).

### ReadXL

The `readxl` package is used for easy load and write data in Microsoft Excel format. To install `readxl` package:
```{r Install readxl package}
install.packages("readxl")
```

For further reference, visit the [readxl website](https://readxl.tidyverse.org) and the [official package documentation](https://cran.r-project.org/web/packages/readxl/).

### Janitor

The `janitor` package has simple functions for examining and cleaning dirty data, saving time on preparing data prior using it. To install `janitor` package:
```{r Install janitor package}
install.packages("janitor")
```

For further reference, visit the [janitor website](http://sfirke.github.io/janitor/) and the [official package documentation](https://cran.r-project.org/web/packages/janitor/index.html).

### SF

The `sf` package is used for enable *simple features*, a standardized way to encode spatial vector data and perform geometrical operations with it through GEOS library. Also allows using the GDAL library to read and write several geographic data formats like Shapefile, GeoPackage, GeoJSON, etc.

To install `sf` package:
```{r Install sf package}
install.packages("sf")
```

**Note for macOS and Linux users**: Before installing the `sf` package, it may be necessary to install the GDAL and PROJ libraries on the system via Terminal. For details on how to do it, please refer to [the instructions provided in the sf website](https://r-spatial.github.io/sf/).

For further reference, visit the [sf website](https://r-spatial.github.io/sf/) and the [official package documentation](https://cran.r-project.org/web/packages/sf/index.html).

### Raster

The `raster` package is used for reading, writing, manipulating, analyzing and modeling spatial data, mainly in raster format. To install `raster` package:
```{r Install raster package}
install.packages("raster")
```

For further reference, visit the [raster website](https://rspatial.org/raster/) and the [official package documentation](https://cran.r-project.org/web/packages/raster/index.html).

### Spatstat

The `spatstat` package is used for spatial statistics and analysing spatial point patterns in 2D (with some support for 3D and very basic support for space-time). To install `spatstat` package:
```{r Install spatstat package}
install.packages("spatstat")
```

For further reference, visit the [spatstat website](http://spatstat.org) and the [official package documentation](https://cran.r-project.org/web/packages/spatstat/index.html).

### TMap

The `tmap` package is used for drawing thematic maps in a similar fashion to `ggplot2` grammar.

To install `tmap` package:
```{r Install tmap package}
install.packages("tmap")
```

For further reference, visit the [tmap website](https://mtennekes.github.io/tmap/) and the [official package documentation](https://cran.r-project.org/web/packages/tmap/index.html).

### Leaflet

The `leaflet` package makes it easy to integrate and control interactive maps in R using the popular *Leaflet* JavaScript library for spatial data visualization.

To install `leaflet` package:
```{r Install leaflet package}
install.packages("leaflet")
```

For further reference, visit the [leaflet website](https://rstudio.github.io/leaflet/) and the [official package documentation](https://cran.r-project.org/web/packages/leaflet/).

### DT (DataTables)

The `DT` package is used to display R dataframes as interactive tables using a JavaScript library named *DataTable*.

To install `DT` package:
```{r Install DT package}
install.packages("DT")
```

For further reference, visit the [DT website](https://rstudio.github.io/DT/) and the [official package documentation](https://cran.r-project.org/web/packages/DT/index.html).

### Complementary packages

It is also possible to install multiple packages at once passing the names as a vector:
```{r Install complementary packages}
install.packages(c("rmapshaper", "geogrid"))
```

Sometimes because of specific reasons it is necessary to install discontinued packages (marked as *archived* in the CRAN). these packages cannot be installed in the regular way as it has been done previously, so it is required to refer to the package's source code and compile it. Later in this workshop we will use a package named `aspace`, now archived and not updated on the CRAN since a lot of years.

In order to compile packages in Windows, it is necessary to install **RTools** from [https://cran.rstudio.com/bin/windows/Rtools/](https://cran.rstudio.com/bin/windows/Rtools/). Once installed, compile and install the package using the `install.packages` function with some extra parameters:
```{r Install aspace package from source}
# First install some dependency packages:
install.packages(c("splancs", "shapefiles"))

# Install aspace from source code:
install.packages("https://cran.r-project.org/src/contrib/Archive/aspace/aspace_3.2.tar.gz",
                 repos = NULL,
                 type = "source")
```

It is possible to install any package from source code using the procedure below presented.

## References

* Xie, Y., Allaire, J. J., & Grolemund, G. (2018). *R markdown: The definitive guide.  [https://bookdown.org/yihui/rmarkdown/](https://bookdown.org/yihui/rmarkdown/)*. CRC Press.
* Wickham, H., & Grolemund, G. (2017). *R for data science: Import, tidy, transform, visualize and model data. [https://r4ds.had.co.nz](https://r4ds.had.co.nz)*. O'Reilly.
* *RStudio Cheatsheets*. (n.d.). Rstudio.com. Retrieved September 8, 2021, from [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/)
* Tennekes, Martijn. (2018). *tmap : Thematic Maps in R*. Journal of Statistical Software. 84. 10.18637/jss.v084.i06, from [https://www.researchgate.net/publication/324652152_tmap_Thematic_Maps_in_R/fulltext/5ad9e7eb0f7e9b28593cf867/tmap-Thematic-Maps-in-R.pdf] (https://www.researchgate.net/publication/324652152_tmap_Thematic_Maps_in_R/fulltext/5ad9e7eb0f7e9b28593cf867/tmap-Thematic-Maps-in-R.pdf)
